Decompositions of congruence subgroups \\ of Chevalley groups

### 2018-01-18 paper.tex
Added affiliation
Purged all comments

%\oddsidemargin 5mm
%\marginparwidth 5mm
%\topmargin 0mm
%\textheight 225mm
%\textwidth 165mm
%\headheight 0mm
%\headsep 10mm
%\footskip 5mm

@@

\title[Decompositions of congruence subgroups of Chevalley groups]{Decompositions of congruence subgroups \\ of Chevalley groups}
\keywords {parabolic decomposition, bounded generation, $\SL_2$-factorization, subsystem factorization, product decomposition. {\em Mathematical Subject Classification (2010):} 20G40, 20G35, 19B14}
\author{Sergey Sinchuk}
\author{Andrei Smolensky}
\email{sinchukss {\it at} yandex.ru; andrei.smolensky {\it at} gmail.com}
\thanks{The authors of the present paper acknowledge the financial support of the Russian Science Foundation grant 14-11-00297}

@@

\section{Introduction}\label{sec:intro}
The aim of this paper is to generalize several decompositions of Chevalley groups over rings 
 to their principal congruence subgroups and relative elementary subgroups.
We focus on the so-called ``parabolic decompositions'', i.e. decompositions involving parabolic and unipotent factors. 
One of our main results is the following theorem.
\begin{thm}[Dennis---Vaserstein decomposition]\label{thm:DennisVaserstein}
Let $\Phi \neq \rE_8$ be a reduced irreducible root system of rank $\ell\geqslant 2$ and let $\{ r, s \}$ be a pair of distinct terminal vertices of the Dynkin diagram of $\Phi$.
Denote the distance between $r$ and $s$ by $d$.
Assume that $\Phi$, $I$ and $\{r, s\}$ satisfy either of the following assumptions:
\begin{thmlist}
 \item $\sr(I) \leqslant d$ for classical $\Phi$,
 \item $\sr(I) \leqslant d$ for $\Phi=\rE_\ell$ with $\{r, s\} = \{ 2, \ell \}$,
 \item $\asr(I) \leqslant d$ for $\Phi=\rE_\ell$ with $\{r, s\} = \{1, \ell \}$,
 \item $\asr(I) \leqslant d$ for $\Phi=\rF_4$.
\end{thmlist}
Then $\E(\Phi, R, I) = \EP_r(R, I) \cdot \U(\Sigma^-_r \cap \Sigma^-_s, I) \cdot \EP_s(R, I).$
\end{thm}
Even in the special case $I=R$ the above theorem contains several new cases as compared to~\cite[Theorems~2.5 and~4.1]{St78}.
%The term ``Dennis---Vaserstein decomposition'' is due to A.~Suslin and M.~Tulenbaev \cite[Lemma~2.1]{ST76}.
The condition $\asr(I) \leq d$ is a relative version of the \emph{absolute stable range} condition of D.~Estes and J.~Ohm, which depends only on the ideal $I$ (see section~\ref{sec:rel-asr}).
Another results of the paper are relative analogues of Gauss and Bass---Kolster decompositions (see~\cref{thm:srRI1,thm:BassKolster}).

We apply our decompositions to several problems. First, we obtain upper estimates of the number of factors in the presentation of a Chevalley group as a product of its subsystem subgroups of type $\rA_\ell$.
For example, in~\cref{thm:SL2width} such estimates are given for relative classical groups and products of subgroups $\SL(2, R, I)$.
% For classical groups over rings satisfying suitable dimension assumptions we show the existence of 
% factorizations with $3|\Phi^+|$ in case $\Phi=\rA_\ell, \rC_\ell$ and $4|\Phi^+|$ otherwise, see
Our another result is the decomposition in terms of subgroups of type $\rA_\ell$ of submaximal rank. 
Using~\cref{thm:DennisVaserstein} we prove the following theorem inspired by the main result of~\cite{Nik07}.
\begin{thm}\label{thm:spin-sln-prod}
Assume that $\sr(I) \leqslant 2$. Then the group $\Epin(2\ell, R, I)=\E(\rD_\ell, R, I)$ is a product of at most $9$ regularly embedded subgroups of type $\rA_{\ell-1}$.
\end{thm}

@@

\section{Preliminaries}\label{sec:preliminaries}
%Let us now fix the notations used in the paper.
Let $\Phi$ be a reduced irreducible root system of rank $\ell$ with a fixed basis of fundamental roots $\Pi=\{\alpha_1, \ldots, \alpha_l\}$.
We use the conventional numbering of basis vectors of $\Pi$ which follows Bourbaki (see~\cite[Table~1]{PSV98}).

@@

%The following recent result of A.~Stepanov served as a starting point for the present paper.
%The notation for the subgroups involved in the statements of the results below is introduced properly in section~\ref{sec:prelim}.
%\begin{prop}[{\cite[Theorem~3.4]{S}}] \label{prop:Stepanov}
%Let $\Phi$ be a reduced irreducible root system of rank $\geq 2$, let $S \subset \Phi$ be an arbitrary parabolic subset of roots, and $\Sigma_S$ be the special part of $S$. The relative elementary subgroup $\E(\Phi, R, I)$ is generated as an abstract group by the set $\mathcal{Z}(\Sigma_S)$. \end{prop}

%The purpose of this paper is two-fold. 
%First, we obtain relative analogues of several classical parabolic factorizations of Chevalley groups such as the Gauss, Bass---Kolster and Dennis---Vaserstein decompositions.
%Then we apply these factorizations to the study of various bounded generation properties of Chevalley groups over rings.
%
%Our first result is a relative version of the Gauss decomposition formulated in terms of the \emph{stable rank} of the ideal $I$ (see section~\ref{sec:stability-conditions} for the definition)
%\begin{thm}\label{thm:srRI1}
%Let $\Phi$ be a root system, let $I$ be an ideal of an arbitrary commutative ring $R$, and assume that $\sr(I)=1$.
%Then the relative elementary Chevalley group $\E(\Phi, R, I)$ admits the following decomposition:
%\[ \E(\Phi, R, I) = \Hh(\Phi, R, I) \cdot \U(\Phi, I) \cdot \U(\Phi^-, I) \cdot \U(\Phi, I). \]
%\end{thm}
%
%The next main result is the relative Dennis---Vaserstein decomposition, which holds for a broader class of rings as compared to \cref{thm:srRI1}.
%\begin{thm}\label{thm:DennisVaserstein}
%Let $\Phi \neq \rE_8$ be a reduced irreducible root system of rank $\ell\geqslant 2$ and let $\{ r, s \}$ be a pair of distinct terminal vertices of the Dynkin diagram of $\Phi$. Denote the distance between $r$ and $s$ by $d$.
%Assume that $\Phi$, $I$ and $\{r, s\}$ satisfy either of the following assumptions:
%\begin{thmlist}
% \item $\sr(I) \leqslant d$ for classical $\Phi$,
% \item $\sr(I) \leqslant d$ for $\Phi=\rE_\ell$ with $\{r, s\} = \{ 2, \ell \}$,
% \item $\asr(I) \leqslant d$ for $\Phi=\rE_\ell$ with $\{r, s\} = \{1, \ell \}$,
% \item $\asr(I) \leqslant d$ for $\Phi=\rF_4$.
%\end{thmlist}
%Then $\E(\Phi, R, I) = \EP_r(R, I) \cdot \U(\Sigma^-_r \cap \Sigma^-_s, I) \cdot \EP_s(R, I).$ \end{thm}
%Even in the special case $I=R$ the above theorem contains several new cases as compared to~\cite[Theorems~2.5 and~4.1]{St78}.
%The term ``Dennis---Vaserstein decomposition'' is due to A.~Suslin and M.~Tulenbaev (cf.~\cite[Lemma~2.1]{ST76}).
%
%The condition $\asr(I) \leq d$ used in the statement of \cref{thm:DennisVaserstein} is a relative version of the \emph{absolute stable range} condition of D.~Estes and J.~Ohm, which depends only on the ideal $I$ (see section~\ref{sec:rel-asr}).
%
%Let $G$ be an abstract group with a generating set $X \subset G$. 
%We denote by $W(G, X)$ the \emph{width of $G$ with respect to $X$}, i.\,e. the smallest natural number $N$ such that every element of $G$ is a product of at most $N$ elements of $X$.
%
%Our first application of the parabolic factorizations is the following theorem, which gives an upper bound on the width of $\E(\Phi, R)$ in terms of Tits generators. 
%Thus, it is a quantitative analogue of \cref{prop:Stepanov}
%and a relative analogue of the results of~\cite{Tavgen91, VseUnitrZ1p, VavSmSuUnitrEng}.
%\begin{thm}\label{thm:width} Let $I$ be an ideal of a commutative ring $R$.
%\begin{thmlist}
%\item If $R=\mathcal{O}_S$ is a Dedekind ring of arithmetic type posessing a real embedding and $\Phi$ is classical of rank $\geqslant2$, then 
%$W(\G(\Phi, R, I), \mathcal{Z}(\Sigma_\ell^-))$ is finite;
%\item If $\sr(I) = 1$ and $\Phi$ is an arbitrary irreducible root system, then 
%\[W(\E(\Phi, R, I), \mathcal{Z}(\Pi))\leqslant 3|\Phi^+|+2\rk(\Phi)-1;\]
%\item If $R = \mathbb{Z}[\sfrac{1}{p}]$ for some prime number $p$, then under the assumption of the Generalized Riemann Hypothesis one has
%\begin{alignat*}{2}
%& W(\E(\Phi, R, I), \mathcal{Z}(\Sigma_\ell^-))\leqslant 3|\Phi^+| + 2\rk(\Phi) + 1 & \text{ if } \Phi=\rA_\ell, \rC_\ell, \\
%& W(\E(\Phi, R, I), \mathcal{Z}(\Sigma_\ell^-))\leqslant 4|\Phi^+| + \rk(\Phi) + 1 & \text{ if } \Phi=\rB_\ell, \rD_\ell.
%\end{alignat*}
%\end{thmlist}
%\end{thm}
%The first assertion of the above theorem is essentially a reformulation of a previously known result of O.~Tavgen (see \cite{TavgenThesis}, cf.~\cite{Mo07}), 
%while for the second and the third assertion we give a direct proof.
%Compare \cref{thm:width} with a recent result of U.~Hadad and D.~W.~Morris \cite[Theorem~1.6]{Ha12} which asserts that 
%$W(\SL(n, \mathbb{Z}, I), \mathcal{Z}(\Pi))$ is finite for any ideal $I \leq \mathbb{Z}$ and $n \geq 3$.
%
%Another application of the parabolic factorizations developed in the present paper is the subsystem factorization in terms of subgroups of type $\rA_1$.
%The case of a finite field $R=F$ was studied by M.~Liebeck, N.~Nikolov and A.~Shalev in~\cite{LNS11}.
%More precisely, for an untwisted Chevalley group it has been proved that $\G(\Phi, F)$ is a product of at most $N=5|\Phi^+|$ copies of $\SL(2, F)$.
%
%In~\cite{V13} N.~Vavilov and E.~Kovach noted that the Bruhat decomposition immediately implies a bound of $N=3|\Phi^+|$ factors for an arbitrary field $F$.
%In fact, in the assumptions of \cref{thm:srRI1} one can immediately obtain the decomposition of $\E(\Phi, R, I)$ with $3|\Phi^+|$ factors of type $\rA_1$.
%Our next result further extends this to an even larger class of rings (which, for example, includes the ideals of all Dedekind rings).
%\begin{thm}\label{thm:SL2width}
%Let $I\trianglelefteq R$ be an ideal, $\Phi$ be an irreducible classical root system satisfying one of the following assumptions.
%\[\begin{array}{l@{\quad}l@{\quad}l}
%\Phi=\rA_\ell, & \sr(I)\leqslant 2, & N=3\left|\Phi^+\right| - \rk(\Phi) - 1; \\
%\Phi=\rC_\ell, & \sr(I)\leqslant 3, & N=3\left|\Phi^+\right| - 2; \\
%\Phi=\rB_\ell, \rD_\ell, & \asr(I)\leqslant 2, & N=4\left|\Phi^+\right| - 3\rk(\Phi).
%\end{array}\]
%Then the principal congruence subgroup $\G(\Phi, R, I)$ can be presented as a product of at most $N$ copies of its (regularly embedded) subgroups isomorphic to $\SL_2(R, I)$.
%\end{thm}
%
%Yet another application of the parabolic factorizations concerns the product decompositions of Chevalley groups formulated in terms of subgroups of type $\rA_\ell$ of submaximal rank.
%Recall that in \cite{Nik07} N.~Nikolov proves the following result. 
%\begin{thm*} Let $G$ be a classical Chevalley group (of normal or twisted type) of rank $\ell$ over a finite field $F$.
%Then $G$ can be presented as a product of at most $200$ conjugates of its subgroups of type $\rA_{\ell-1}$. \end{thm*}
%
%It is classically known that over an arbitrary field $F$ the group $\SL_{n+1}(F)$, $n\geq 3$ is a product of at most $4$ subgroups of type $\SL(n, F)$.
%On the other hand, one immediately obtains from \cref{thm:DennisVaserstein} and surjective stability for linear $\K_1$-functor that $\SL_{n+1}(R, I)$ factors into a product
%of a most $5$ subgroups of type $\SL_n(R, I)$ for any ideal $I$ satisfying $\sr(I)\leqslant n-1$.
%
%Applying \cref{thm:DennisVaserstein} and proving a more accurate version of \cite[Proposition~1]{Nik07} in the special case $\Phi=\rD_\ell$, we obtain the following result.
%\begin{thm}\label{thm:spin-sln-prod}
%Assume that $\sr(I) \leqslant 2$. Then the group $\Epin_{2\ell}(R, I)=\E(\rD_\ell, R, I)$ is a product of at most $9$ conjugates of its (regularly embedded) subgroups of type $\rA_{\ell-1}$.
%\end{thm}

%The rest of the article is organized as follows.
%In sections \ref{sec:prelim}--\ref{sec:stability-conditions} we introduce the principal notation and prove some basic facts pertaining to Chevalley groups over rings and the stability conditions.
%In section \ref{sec:factorizations} we construct various parabolic factorizations and thus prove \cref{thm:srRI1,thm:DennisVaserstein,thm:SL2width}.
%Finally, we treat the bounded generation and the product decompositions separately in section~\ref{sec:applications}, where \cref{thm:width,thm:spin-sln-prod} are proved.

@@

Define the semisimple root elements $h_\alpha(\varepsilon)$, $\alpha\in\Phi$, $\varepsilon\in R^*$ as $h_\alpha(\varepsilon)=w_\alpha(\varepsilon)w_\alpha(-1)$.
The elements $h_\alpha(\varepsilon)$ satisfy the following relation:
\begin{alignat}{2} 
%& [ h_\alpha(\varepsilon_1), h_\beta(\varepsilon_2)] = 1, &\quad& \alpha, \beta \in \Phi, \ \varepsilon_1, \varepsilon_2 \in R^*; \label{rel:h-comm} \\ 
%& \phantom{[} h_\alpha(\varepsilon_1 \varepsilon_2) = h_\alpha(\varepsilon_1) h_\alpha(\varepsilon_2), && \alpha \in \Phi, \ \varepsilon_1, \varepsilon_2 \in R^*; \label{rel:h-mult} \\
& \phantom{[} h_\alpha(\varepsilon)x_\beta(\xi)h_\alpha(\varepsilon)^{-1} = x_\beta\left(\varepsilon^{\langle\beta, \alpha\rangle}\xi\right), && \alpha, \beta \in \Phi, \ \varepsilon \in R^*, \xi \in R. \label{rel:h-w}
\end{alignat}

%Our notation and terminology pertaining to the representations of Chevalley groups follows \cite[\S~1.4]{PSV98} and \cite[\S~I.2]{Ma69})
%Let $\pi$ be an irreducible representation of $\G(\Phi, R)$ acting on a free $R$-module $V$.
%We denote by $\Lambda=\Lambda(\pi)$ the set of weights of $\pi$ and by $\Lambda(\pi)^*$ the subset of non-zero weights.
%Denote by $V_\lambda \leq V$ the \emph{weight subspace}, corresponding to $\lambda\in\Lambda(\pi)$ and denote by $m_\lambda$ the dimension of $V_\lambda$.
%
%A representation $\pi$ is called \emph{basic} if for every pair of non-zero weights $\lambda, \mu \in \Lambda^*(\pi)$ such that $\lambda - \mu \in \Pi$ one has $\sigma_\alpha(\lambda) = \mu$ for some $\alpha\in\Phi$.
%Obviously, this is equivalent to saying that $W(\Phi)$ acts transitively on $\Lambda(\pi)^*$.

%The irreducible representations of Chevalley groups used in the present paper form a sublist of~\cite[Table~2]{PSV98}.
%More specifically, all of them are \emph{fundamental} (i.\,e. their highest weight is fundamental) and basic.
%We widely employ the technique of \emph{weight diagrams} (see~\cite[\S~2]{PSV98}) to help visualize the structure of these representations.
%In the simplest case of a \emph{microweight} representation $\pi$ (i.\,e. $\pi$ is basic and $V_0 = 0$)
%the weight diagram of $(V, \pi)$ is just an undirected graph whose vertices are in the one-to-one correspondence with the elements of $\Lambda(\pi)$.
%Two vertices $\lambda$, $\mu$ of the diagram are joined by a bond marked $i$ if $\lambda-\mu = \alpha_i$.
%
%An immediate application of the weight diagrams is that they allow one to quickly read off the branching of a given basic representation with respect to a subsystem subgroup.
%If $J$ is a subset of $\Pi$ and $\Delta$ is the subsystem of $\Phi$ spanned by $J$ then the weight diagram of $\pi$ restricted to $G(\Delta, R)$ can be obtained
%from the weight diagram of $\pi$ by simply removing all the bonds whose label $i$ is such that $\alpha_i\not\in J$ (cf.~\cite[\S~2.7]{PSV98}).
% 
%Recall that all nonzero weights of a basic representation have multiplicty one and the multiplicity $m_0$ of the zero weight subspace equals $|\Delta(\pi)|$, where $\Delta(\pi)= \Lambda(\pi)^* \cap \Pi$ (see~\cite[Lemma~2.1]{Ma69}).
%For $\alpha \in \Phi$ one can define certain elements $e_0(\alpha) \in V_0$ and $\alpha_* \in V_0^* = \Hom_R(V_0, R)$
%and choose some set of vectors $E^* = \{ e_\lambda \}_{\lambda \in \Lambda^*(\pi)}$ in such a way that 
%$E_\pi = E^* \cup \{ e_0(\alpha) \}_{\alpha \in \Delta(\pi)}$ forms a basis of $V$ in which the action of the elementary root unipotents
%$x_\alpha(\xi)$ has a particularly simple description.
%
%\begin{lemma}[{\cite[Lemma~2.3]{Ma69}}]\label{lemma:Matsumoto}
%Let $(V, \pi)$ be a basic representation of $\G(\Phi, R)$.
%Then the action of $x_\alpha(\xi)$ is described by the following formulae.
%\newline \begin{tabular}{ll}
%If $\lambda\in\Lambda(\pi)^*, \ \lambda+\alpha\notin\Lambda(\pi)$, then & $x_\alpha(\xi)e_\lambda=e_\lambda$; \\
%If $\lambda, \lambda+\alpha\in\Lambda(\pi)^*$, then & $x_\alpha(\xi)e_\lambda=e_\lambda\pm\xi e_{\lambda+\alpha}$; \\
%If $\alpha\notin\Lambda(\pi)^*$, then & $x_\alpha(\xi)v_0=v_0$ for any $v_0\in V_0$; \\
%If $\alpha\in\Lambda(\pi)^*$, then & $x_\alpha(\xi)e_{-\alpha} = e_{-\alpha}\pm\xi e_0(\alpha)\pm\xi^2 e_\alpha$ and \\
%                                   & $x_\alpha(\xi)v_0         = v_0\pm\xi\alpha_*(v_0)e_\alpha$ for $v_0 \in V_0$.
%\end{tabular} \end{lemma}

%\subsection{Relative Chevalley groups}\label{sec:relative-elementary}
%Recall from the introduction that the elements $z_\alpha(s, \xi) = x_\alpha(s)^{x\ssub{-\alpha}(\xi)}$ generate $\E(\Phi, R, I)$ as an abstract group.

@@

%\subsection{\texorpdfstring{$\K_1$}{K1}-functor modeled on Chevalley groups}
%Recall that by a theorem of Taddei $\E(\Phi, R)$ is a normal subgroup of $\G_{sc}(\Phi, R)$ provided $\Phi$ is an irreducible root system of rank $\geqslant 2$ (see~\cite{Ta}).
%
%Using standard relativization argument it is not hard to deduce from Taddei's result that $\E(\Phi, R, I)$ is normal in $\G(\Phi, R, I)$ under the same assumptions on $\Phi$.
%This allows us to define the relative $\K_1$-group as $\K_1(\Phi, R, I)=\G(\Phi, R, I)/\E(\Phi, R, I).$
%When $I=R$, we write $\K_1(\Phi, R)$ for $\K_1(\Phi, R, R)$.
%
%In some cases it is known that $\K_1(\Phi, R, I)$ is trivial.
%For example, $\SK_1(\ell+1, R)=\K(\rA_\ell, R)=1$ for any ring of stable rank $1$ (see section~\ref{sec:stability-conditions}), 
%and for other root systems some stronger assumption is required such as $\asr(R)=1$ or being semilocal. 
%For any Euclidean ring $\K_1(\Phi, R, I)$ is trivial for every root system $\Phi$.
%
%Let $k$ be a global field, $S$ be a set of places of $k$, 
%$\mathcal{O}_S$ be the Dedekind ring of arithmetic type defined by $S$ and $I$ be an ideal of $\mathcal{O}_S$.
%\begin{prop} \label{prop:K1triv}
%Let $\Phi$ be a root system of rank $\geqslant2$. Assume that the field $k$ has a real embedding. Then $\K_1(\Phi, \mathcal{O}_S, I)=1$.
%\end{prop}
%\begin{proof} Follows from \cite[Theorem~3.6]{BMS} and \cite[Corollary~4.5]{Ma69}. \end{proof}

@@

%The following proposition summarizes the basic properties of the stable rank.
%\begin{prop} \label{prop:sr_properties}
%Let $R$ be an arbitrary commutative unital ring and let $I\trianglelefteq R$ be an ideal.
%\begin{lemlist}
%\item For any ideal $I'\subseteq I$ one has $\sr(I')\leqslant\sr(I)$, $\sr(I/I')\leqslant\sr(I)$;
%\item One has $\sr\Bigl(\prod\limits_{i=1}^n R_i\Bigr)=\max\limits_{i=1}^n\bigl(\sr(R_i)\bigr)$;
%\item Let $J(R)$ denote the Jacobson radical of $R$, then $\sr(R)=\sr(R/J(R))$ and $\sr(J(R))=1$;
%\item Let $I\rtimes\mathbb{Z}$ denote the ring obtained by formally adjoining $1$ to $I$. Then $\sr(I\rtimes\mathbb{Z})=\max(2,\sr(I))$.
%\end{lemlist}
%\end{prop}
%\begin{proof} See~\cite[Theorem~2.3]{Va69}, \cite{Va71}. 
%\end{proof}
%\begin{example}
%Since the stable rank of a field equals $1$, one can conclude from the previous proposition that $\sr(R)=1$ for any semilocal ring $R$.
%Other examples of rings of stable rank $1$ are provided by the Boolean rings, the ring of all algebraic integers, the ring of entire functions, the disk-algebra, etc. For this and many other examples see~\cite{VasSR1}.
%
%%Formally adjoining a unit to a non-unital ring of stable rank $1$ produces a non-trivial example of an ideal having the stable rank strictly lower that the ambient ring.
%\end{example}

@@

Let $R$ be a commutative ring. We denote by $\Max(R)$ its \emph{maximum spectrum}, i.e. the set of maximal ideals of $R$, equipped with the Zariski topology. For a topological space $X$ denote by $\dim(X)$ its usual topological dimension.
From the definition of $\asr(I)$ and \cite[Theorem~2.3]{EO} (or~\cite[Theorem~3.7]{MKV}) it follows that
\begin{equation} \label{sr-estimates} \sr(I)\leqslant\asr(I)\leqslant\asr(R)\leqslant \dim(\Max(R))+1\leqslant\dim(\Spec(R))+1. \end{equation}
%There exist examples of rings for which the first inequality in the above formula is strict.
%More precisely, there are rings of stable rank $1$ but of absolute stable rank $\geqslant2$, see~\cite{MKV}.
%On the other hand, \cite[Theorem~1.3]{MKV} shows that $\sr(R)=\asr(R)$ for any principal ideal ring $R$.
%\begin{example}
%From inequality~\eqref{sr-estimates} one concludes, in particular, that 
%\begin{itemize}
%\item $\asr(R) = 1$ for every semilocal ring $R$;
%\item $\asr(R)\leqslant2$ for every Dedekind domain $R$;
%\item $\asr(A[x])\leqslant3$ for every locally principal ring $A$;
%\item $\asr(F[x_1,\ldots,x_n])\leqslant n+1$ for any field $F$.
%\end{itemize}
%For certain classes of fields there is also a lower estimate for $\sr(F[x_1, \ldots, x_n])$ which allows to compute precisely the value of both $\sr(R)$ and $\asr(R)$.
%For example, $\sr(F[x,y])=3$ for every $F$ with $\K_2^M(F)\neq0$ (see~\cite{Kru73}) and $\sr(F[x_1, \ldots, x_n])=n+1$ for $F\subseteq\mathbb{R}$, see~\cite[Theorem~8]{Va71}.
%Notice also that $\sr(\mathbb{Z}[x])=3$ (see \cite{GMV}) and $\sr(\mathbb{Z}[x_1,\ldots,x_n])=n+1$ for $n>1$ (see example after~\cite[Theorem~18.2]{VS76}).
%\end{example}

@@

Now set $X=\left\{x_\alpha(\xi)\ \middle|\ \alpha\in\Pi, \ \xi\in I \right\}$. 
Every root is a conjugate of some fundamental root under the action of $W(\Phi)$.
Since $\E(\Phi, R)$ contains the normalizer of the torus, the set $X^{\E(\Phi, R)}$ contains all the generators of the group $\E(\Phi, I) = \langle x_\alpha(\xi), \  \alpha\in\Phi, \ \xi\in I \rangle$.
Finally, the inclusion $XY \subseteq Y$ follows from the fact that $\Hh(\Phi, R, I)$ normalizes every root subgroup $X_\alpha(I)$.
\end{proof}
\begin{thm}\label{thm:srRI1}
Let $\Phi$ be a root system, let $I$ be an ideal of an arbitrary commutative ring $R$, and assume that $\sr(I)=1$.
Then the relative elementary Chevalley group $\E(\Phi, R, I)$ admits the following decomposition:
\[ \E(\Phi, R, I) = \Hh(\Phi, R, I) \cdot \U(\Phi, I) \cdot \U(\Phi^-, I) \cdot \U(\Phi, I). \]
\end{thm}
\begin{proof} %[Proof of \cref{thm:srRI1}]
In view of \cref{thm:Gauss} it suffices to prove that the Gauss decomposition of length $3$ holds for $\Phi=\rA_1$. Let $A=\begin{psmallmatrix}a & b \\ c & d\end{psmallmatrix}$ be an element of $\SL(2, R, I)$.
The first column of $A$ is $I$-unimodular, therefore there exists $z\in I$ such that $a+cz\in R^*$.
Multiplying $A$ on the left by $x_{12}(z)$, we get a matrix $A'=x_{12}(z)\cdot A=\begin{psmallmatrix}a' & b' \\ c & d\end{psmallmatrix}$ with invertible element $a'$ in the top-left corner.
After multiplying $A'$ on the left by $x_{21}(-c/a')$ and on the right by $x_{12}(-b'/a')$ we get a diagonal matrix. 
Thus we have obtained the relative Gauss decomposition of $A$
\begin{equation*}
A=x_{12}(-z)\cdot x_{21}(c/a')\cdot
\begin{pmatrix} \varepsilon & 0 \\ 0 & 1/\varepsilon \end{pmatrix}
\cdot x_{12}(b'/a')=x_{12}(-z)\cdot
\begin{pmatrix} \varepsilon & 0 \\ 0 & 1/\varepsilon \end{pmatrix}
\cdot x_{21}(y) \cdot x_{12}(b'/a'), 
\end{equation*}
where $\varepsilon\in 1+I$ and $y\in I$. \end{proof}

>>>>

@@

\title[Decompositions of congruence subgroups of Chevalley groups]{Decompositions of congruence subgroups \\ of Chevalley groups}
\keywords {parabolic decomposition, bounded generation, $\SL_2$-factorization, subsystem factorization, product decomposition. {\em Mathematical Subject Classification (2010):} 20G40, 20G35, 19B14}
\author{Sergey Sinchuk}
\address{Chebyshev laboratory, St. Petersburg State University, St. Petersburg, Russia}
\email{sinchukss {\it at} yandex.ru;}

\author{Andrei Smolensky}
\address{Department of Mathematics and Mechanics, St. Petersburg State University, St. Petersburg, Russia}
\email{ andrei.smolensky {\it at} gmail.com}
\thanks{The authors of the present paper acknowledge the financial support of the Russian Science Foundation grant 14-11-00297}

@@

\section{Introduction}\label{sec:intro}
The aim of this paper is to generalize several decompositions of Chevalley groups over rings 
 to their principal congruence subgroups and relative elementary subgroups.
We focus on the so-called ``parabolic decompositions'', i.e. decompositions involving parabolic and unipotent factors. 
One of our main results is the following theorem.
\begin{thm}[Dennis---Vaserstein decomposition]\label{thm:DennisVaserstein}
Let $\Phi \neq \rE_8$ be a reduced irreducible root system of rank $\ell\geqslant 2$ and let $\{ r, s \}$ be a pair of distinct terminal vertices of the Dynkin diagram of $\Phi$.
Denote the distance between $r$ and $s$ by $d$.
Assume that $\Phi$, $I$ and $\{r, s\}$ satisfy either of the following assumptions:
\begin{thmlist}
 \item $\sr(I) \leqslant d$ for classical $\Phi$,
 \item $\sr(I) \leqslant d$ for $\Phi=\rE_\ell$ with $\{r, s\} = \{ 2, \ell \}$,
 \item $\asr(I) \leqslant d$ for $\Phi=\rE_\ell$ with $\{r, s\} = \{1, \ell \}$,
 \item $\asr(I) \leqslant d$ for $\Phi=\rF_4$.
\end{thmlist}
Then $\E(\Phi, R, I) = \EP_r(R, I) \cdot \U(\Sigma^-_r \cap \Sigma^-_s, I) \cdot \EP_s(R, I).$
\end{thm}
Even in the special case $I=R$ the above theorem contains several new cases as compared to~\cite[Theorems~2.5 and~4.1]{St78}.
The condition $\asr(I) \leq d$ is a relative version of the \emph{absolute stable range} condition of D.~Estes and J.~Ohm, which depends only on the ideal $I$ (see section~\ref{sec:rel-asr}).
Another results of the paper are relative analogues of Gauss and Bass---Kolster decompositions (see~\cref{thm:srRI1,thm:BassKolster}).

We apply our decompositions to several problems. First, we obtain upper estimates of the number of factors in the presentation of a Chevalley group as a product of its subsystem subgroups of type $\rA_\ell$.
For example, in~\cref{thm:SL2width} such estimates are given for relative classical groups and products of subgroups $\SL(2, R, I)$.
Our another result is the decomposition in terms of subgroups of type $\rA_\ell$ of submaximal rank. 
Using~\cref{thm:DennisVaserstein} we prove the following theorem inspired by the main result of~\cite{Nik07}.
\begin{thm}\label{thm:spin-sln-prod}
Assume that $\sr(I) \leqslant 2$. Then the group $\Epin(2\ell, R, I)=\E(\rD_\ell, R, I)$ is a product of at most $9$ regularly embedded subgroups of type $\rA_{\ell-1}$.
\end{thm}

@@

\section{Preliminaries}\label{sec:preliminaries}
Let $\Phi$ be a reduced irreducible root system of rank $\ell$ with a fixed basis of fundamental roots $\Pi=\{\alpha_1, \ldots, \alpha_l\}$.
We use the conventional numbering of basis vectors of $\Pi$ which follows Bourbaki (see~\cite[Table~1]{PSV98}).

@@

@@

Define the semisimple root elements $h_\alpha(\varepsilon)$, $\alpha\in\Phi$, $\varepsilon\in R^*$ as $h_\alpha(\varepsilon)=w_\alpha(\varepsilon)w_\alpha(-1)$.
The elements $h_\alpha(\varepsilon)$ satisfy the following relation:
\begin{alignat}{2} 
& \phantom{[} h_\alpha(\varepsilon)x_\beta(\xi)h_\alpha(\varepsilon)^{-1} = x_\beta\left(\varepsilon^{\langle\beta, \alpha\rangle}\xi\right), && \alpha, \beta \in \Phi, \ \varepsilon \in R^*, \xi \in R. \label{rel:h-w}
\end{alignat}

@@

@@

@@

Let $R$ be a commutative ring. We denote by $\Max(R)$ its \emph{maximum spectrum}, i.e. the set of maximal ideals of $R$, equipped with the Zariski topology. For a topological space $X$ denote by $\dim(X)$ its usual topological dimension.
From the definition of $\asr(I)$ and \cite[Theorem~2.3]{EO} (or~\cite[Theorem~3.7]{MKV}) it follows that
\begin{equation} \label{sr-estimates} \sr(I)\leqslant\asr(I)\leqslant\asr(R)\leqslant \dim(\Max(R))+1\leqslant\dim(\Spec(R))+1. \end{equation}

@@

Now set $X=\left\{x_\alpha(\xi)\ \middle|\ \alpha\in\Pi, \ \xi\in I \right\}$. 
Every root is a conjugate of some fundamental root under the action of $W(\Phi)$.
Since $\E(\Phi, R)$ contains the normalizer of the torus, the set $X^{\E(\Phi, R)}$ contains all the generators of the group $\E(\Phi, I) = \langle x_\alpha(\xi), \  \alpha\in\Phi, \ \xi\in I \rangle$.
Finally, the inclusion $XY \subseteq Y$ follows from the fact that $\Hh(\Phi, R, I)$ normalizes every root subgroup $X_\alpha(I)$.
\end{proof}
\begin{thm}\label{thm:srRI1}
Let $\Phi$ be a root system, let $I$ be an ideal of an arbitrary commutative ring $R$, and assume that $\sr(I)=1$.
Then the relative elementary Chevalley group $\E(\Phi, R, I)$ admits the following decomposition:
\[ \E(\Phi, R, I) = \Hh(\Phi, R, I) \cdot \U(\Phi, I) \cdot \U(\Phi^-, I) \cdot \U(\Phi, I). \]
\end{thm}
\begin{proof}
In view of \cref{thm:Gauss} it suffices to prove that the Gauss decomposition of length $3$ holds for $\Phi=\rA_1$. Let $A=\begin{psmallmatrix}a & b \\ c & d\end{psmallmatrix}$ be an element of $\SL(2, R, I)$.
The first column of $A$ is $I$-unimodular, therefore there exists $z\in I$ such that $a+cz\in R^*$.
Multiplying $A$ on the left by $x_{12}(z)$, we get a matrix $A'=x_{12}(z)\cdot A=\begin{psmallmatrix}a' & b' \\ c & d\end{psmallmatrix}$ with invertible element $a'$ in the top-left corner.
After multiplying $A'$ on the left by $x_{21}(-c/a')$ and on the right by $x_{12}(-b'/a')$ we get a diagonal matrix. 
Thus we have obtained the relative Gauss decomposition of $A$
\begin{equation*}
A=x_{12}(-z)\cdot x_{21}(c/a')\cdot
\begin{pmatrix} \varepsilon & 0 \\ 0 & 1/\varepsilon \end{pmatrix}
\cdot x_{12}(b'/a')=x_{12}(-z)\cdot
\begin{pmatrix} \varepsilon & 0 \\ 0 & 1/\varepsilon \end{pmatrix}
\cdot x_{21}(y) \cdot x_{12}(b'/a'), 
\end{equation*}
where $\varepsilon\in 1+I$ and $y\in I$. \end{proof}
